Bayesian calibration of a hybrid nuclear collision model\\ using $\boldsymbol p$-Pb and Pb-Pb data at energies available\\ at the CERN Large Hadron Collider

### 2019-10-27 paper/paper.tex
Finish referee revisions

\begin{table}[t]
  \caption{
    \label{tab:post_param}
    Posterior parameter estimates corresponding to Fig.~\ref{fig:posterior}.
    The reported values are for the distribution median and 90\% highest posterior density credible interval.
  }
  \begin{ruledtabular}
    \newlength{\cellwidth}
    \settowidth{\cellwidth}{$-0.00$}
    \newcommand{\est}[3]{\parbox{\cellwidth}{\hfill$#1$}$_{-#2}^{+#3}$}
    \begin{tabular}{llll}
      \toprule
      \multicolumn{2}{c}{Initial condition / Pre-eq}     & \multicolumn{2}{c}{QGP medium}              \\
      \cmidrule(r){1-2}                                    \cmidrule(l){3-4}
      \addlinespace[.4ex]
      Norm       & \est{20.0}{2.5}{2.6} GeV      & $\smin$     & \est{0.08}{0.07}{0.07}            \\[1.1ex]
      $p$        & \est{0.002}{0.181}{0.156}     & $\sslope$   & \est{1.23}{1.23}{1.46} GeV$^{-1}$ \\[1.1ex]
      $\sigmaf$  & \est{0.91}{0.33}{0.32}        & $\scrv$     & \est{-0.08}{0.92}{0.79}           \\[1.1ex]
      $\rc$      & \est{0.88}{0.23}{0.26} fm     & $\bmax$     & \est{0.026}{0.026}{0.032}         \\[1.1ex]
      $\nc$      & \est{6.0}{3.4}{3.0}           & $\bwidth$   & \est{0.036}{0.036}{0.043} GeV     \\[1.1ex]
      $\wc$      & \est{0.52}{0.20}{0.28} fm     & $\bloc$     & \est{0.174}{0.024}{0.020} GeV     \\[1.1ex]
      $\dmin$    & \est{1.12}{0.49}{0.58} fm     & $\Tsw$      & \est{0.149}{0.014}{0.013} GeV     \\[1.1ex]
      $\taufs$   & \est{0.47}{0.37}{0.55} \fmc   & \\
      \addlinespace[.4ex]
      \bottomrule
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

@@

\noindent
We also note that although the $p$-Pb posterior samples have a somewhat larger spread than the Pb-Pb samples, the percentage uncertainty of the $p$-Pb emulator is similar to that of the Pb-Pb emulator, and thus the difference is likely due to the larger variance of the $p$-Pb training data.
The uncertainty in the posterior distribution could thus be improved by running the calibration with more design points or with a narrower range of parameter values to increase the density of the training points and reduce interpolation uncertainty.

We now direct our attention to Fig.~\ref{fig:posterior} which shows the main result of this work, the posterior distribution of the model input parameters.
Recall that the posterior $P(\xv | E)$ is the probability that our hypothesis $\xv = \xv_\star$ is correct, given the evidence $E$ provided by the predictions of the model and the experimental data.
The present posterior has 15 dimensions, one for each parameter listed in Table~\ref{tab:design}, and thus its joint distribution cannot be summarized by one figure alone.
We therefore sample the distribution and histogram the samples to project the distribution onto one or two dimensions at a time.
Each diagonal panel is the distribution of a \emph{single} model parameter (marginalized over all others), and each lower-diagonal panel is the joint distribution of a \emph{pair} of model parameters (marginalized over all others).
We also report numeric estimates for each parameter's median value and 90\% credible interval and annotate their values along the distribution diagonal (see Table~\ref{tab:post_param}).
For example, the fictitious parameter estimate $q=2.45_{-0.15}^{+0.20}$ reports a median value $\tilde{q}=2.45$ and 90\% credible interval $2.30 < q < 2.65$.

@@

\begin{figure}[b]
  \includegraphics{inelasticity_profile}
  \caption{
    \label{fig:inelasticity_profile}
    Glauber model estimates of the normalized proton-proton inelasticity density $G_\mathrm{inel}(b) = d^2\sigma_\mathrm{inel}/d^2b$ at two different LHC beam energies.
    Blue line: A multi-source Glauber model fit to LHC data at $\sqrts=7$~TeV \cite{Albacete:2016pmp}.
    Orange line: \trento\ model predictions at $\sqrts=5.02$~TeV using median parameters from the present study's Bayesian posterior.
  }
\end{figure}

Additional constraints on the proton size and shape are provided by the proton-proton inelasticity density $G_\mathrm{inel}(b) = d^2\sigma_\mathrm{inel}(b)/d^2b$, which measures the proton-proton inelastic collision probability as a function of its impact parameter.
We compare, in Fig~\ref{fig:inelasticity_profile}, the normalized proton-proton inelasticity density $G_\mathrm{inel}(b) / G_\mathrm{inel}(0)$ at $\sqrts=5.02$~TeV predicted by \trento\ using the present study's posterior median parameters with a calculation from reference \cite{Albacete:2016pmp}, which extracted the proton-proton inelasticity density from LHC data at $\sqrts=7$~TeV using a multi-source Glauber model \cite{Albacete:2016pmp}.
Our Bayesian median estimate (orange line) prefers a larger width for the proton-proton normalized inelasticity density compared to that of reference \cite{Albacete:2016pmp}, which is \emph{opposite} the trend one would expect based on the differences in beam energies alone.
This suggests that our posterior estimate for the proton radius is somewhat oversized.
Nevertheless, it is fascinating that the present Bayesian estimate is as close as it is, given that the model is fit to quantities which are not typically used to extract the proton-proton inelasticity density.

@@

Second, we obtained a posterior distribution for the model input parameters in Fig.~\ref{fig:posterior} and reported quantitative estimates for these parameters in Table~\ref{tab:post_param}.
Here we summarize our key findings about the model:
\begin{enumerate}[itemsep=0pt, leftmargin=2\parindent]
  \item
    Using \trento\ initial conditions followed by a pre-equilibrium free streaming stage, we find that the initially produced energy density scales like the geometric mean of participant nuclear thickness; see Eq.~\eqref{eq:geometric_mean}.
    Previous Bayesian studies have reported similar scaling for the initially produced \emph{entropy} density \cite{Bernhard:2016tnd, Ke:2016jrd}, although these studies assumed static initialization, an approximation which neglects the initial values of $u^\mu$, $\pi^{\mu\nu}$ and $\Pi$ at the hydrodynamic starting time.
  \item
    We find that nucleon substructure is necessary to simultaneously describe $p$-Pb and Pb-Pb bulk observables.
    However, we observe no strong preference for a specific number of constituents $\nc$ (hot spots) inside the nucleon.
    In particular, we find no evidence to support the specific number $\nc=3$ which is commonly used in the literature.
  \item
    The present Bayesian analysis prefers larger nucleons, $w \sim 1$~fm, in agreement with a similar Bayesian analysis calibrated to Pb-Pb data at $\sqrts=2.76$ and 5.02~TeV using a hybrid model without nucleon substructure \cite{Bernhard:2018hnz}.
    We note that our result is in significant tension with an estimate for the effective nucleon width based on the gluon distribution extracted from HERA data \cite{Rezaeian:2012ji}.
  \item
    We obtain an estimate $\wc=0.52_{-0.20}^{+0.28}$~fm for the Gaussian width of the constituent hot spots inside each nucleon.
    This width is considerably larger than the length scales typically associated with nucleon substructure.
    It may help explain the seemingly unreasonable success of hydrodynamics in small collision systems.
  \item
    We report an estimate $\taufs=0.47_{-0.37}^{+0.55}~\fmc$ for the model's pre-equilibrium free streaming time.
    This is significantly shorter than the estimate obtained from a similar Bayesian study in Ref.~\cite{Bernhard:2018hnz}, which found $\taufs=1.16_{-0.25}^{+0.29}~\fmc$.
    It is not clear whether the difference is a result of nucleon substructure or the different observables used to calibrate each analysis.
  \item
    We compare in Fig.~\ref{fig:region_shear_bulk} our estimate for the temperature dependence of the QGP specific shear and bulk viscosities to those of Ref.~\cite{Bernhard:2018hnz}, which performed a Bayesian calibration to Pb-Pb bulk observables at $\sqrts=2.76$ and 5.02~TeV using a physics model without nucleon substructure.
    The two studies are in good agreement, although Ref.~\cite{Bernhard:2018hnz} obtains a more precise estimate for $(\eta/s)(T)$, likely due to the additional beam energies and observables included and an enhanced sensitivity of larger collision systems to the QGP viscosity.
  \item
    We make predictions in Figs.~\ref{fig:obs_map} and \ref{fig:flow_corr} for several quantities which were not included in the model calibration, including the identified yields, transverse energy, symmetric cumulants, and mean $p_T$ fluctuations at $\sqrts=5.02$~TeV.
    Interestingly, our MAP estimate for the Pb-Pb symmetric cumulants at $\sqrts=5.02$~TeV are significantly different than those estimated in Ref.~\cite{Bernhard:2018hnz}.
    This could be a direct (or indirect) result of including nucleon substructure in the model calibration.
\end{enumerate}

@@

Similarly, the results would greatly benefit from additional beam energies and collision systems.
Notably absent is data from RHIC which includes measurements for $p$-Au, $d$-Au, $^3$He-Au, Cu-Cu, Cu-Au, Au-Au, and \mbox{U-U} collision systems at $\sqrts=200$~GeV.
This data would undoubtedly help constrain the model parameters and would enable more stringent tests of the calibrated model predictions.
The RHIC data may also help elucidate the beam-energy dependence of the model parameters which would be worth investigating.
We leave these improvements for future studies.

>>>>

@@

We also note that although the $p$-Pb posterior samples have a somewhat larger spread than the Pb-Pb samples, the percentage uncertainty of the $p$-Pb emulator is similar to that of the Pb-Pb emulator, and thus the difference is likely due to the larger variance of the $p$-Pb training data.
The uncertainty in the posterior distribution could thus be improved by running the calibration with more design points or with a narrower range of parameter values to increase the density of the training points and reduce interpolation uncertainty.

We now direct our attention to Fig.~\ref{fig:posterior} which shows the main result of this work, the posterior distribution of the model input parameters.
Recall that the posterior $P(\xv | E)$ is the probability that our hypothesis $\xv = \xv_\star$ is correct, given the evidence $E$ provided by the predictions of the model and the experimental data.
The present posterior has 15 dimensions, one for each parameter listed in Table~\ref{tab:design}, and thus its joint distribution cannot be summarized by one figure alone.
We therefore sample the distribution and histogram the samples to project the distribution onto one or two dimensions at a time.

@@

Each diagonal panel is the distribution of a \emph{single} model parameter (marginalized over all others), and each lower-diagonal panel is the joint distribution of a \emph{pair} of model parameters (marginalized over all others).
We also report numeric estimates for each parameter's median value and 90\% credible interval and annotate their values along the distribution diagonal (see Table~\ref{tab:post_param}).
For example, the fictitious parameter estimate $q=2.45_{-0.15}^{+0.20}$ reports a median value $\tilde{q}=2.45$ and 90\% credible interval $2.30 < q < 2.65$.

\begin{table}
  \caption{
    \label{tab:post_param}
    Posterior parameter estimates corresponding to Fig.~\ref{fig:posterior}.
    The reported values are for the distribution median and 90\% highest posterior density credible interval.
  }
  \begin{ruledtabular}
    \newlength{\cellwidth}
    \settowidth{\cellwidth}{$-0.00$}
    \newcommand{\est}[3]{\parbox{\cellwidth}{\hfill$#1$}$_{-#2}^{+#3}$}
    \begin{tabular}{llll}
      \toprule
      \multicolumn{2}{c}{Initial condition / Pre-eq}     & \multicolumn{2}{c}{QGP medium}              \\
      \cmidrule(r){1-2}                                    \cmidrule(l){3-4}
      \addlinespace[.4ex]
      Norm       & \est{20.0}{2.5}{2.6} GeV      & $\smin$     & \est{0.08}{0.07}{0.07}            \\[1.1ex]
      $p$        & \est{0.002}{0.181}{0.156}     & $\sslope$   & \est{1.23}{1.23}{1.46} GeV$^{-1}$ \\[1.1ex]
      $\sigmaf$  & \est{0.91}{0.33}{0.32}        & $\scrv$     & \est{-0.08}{0.92}{0.79}           \\[1.1ex]
      $\rc$      & \est{0.88}{0.23}{0.26} fm     & $\bmax$     & \est{0.026}{0.026}{0.032}         \\[1.1ex]
      $\nc$      & \est{6.0}{3.4}{3.0}           & $\bwidth$   & \est{0.036}{0.036}{0.043} GeV     \\[1.1ex]
      $\wc$      & \est{0.52}{0.20}{0.28} fm     & $\bloc$     & \est{0.174}{0.024}{0.020} GeV     \\[1.1ex]
      $\dmin$    & \est{1.12}{0.49}{0.58} fm     & $\Tsw$      & \est{0.149}{0.014}{0.013} GeV     \\[1.1ex]
      $\taufs$   & \est{0.47}{0.37}{0.55} \fmc   & \\
      \addlinespace[.4ex]
      \bottomrule
    \end{tabular}
  \end{ruledtabular}
\end{table}

@@

\begin{figure}[b]
  \includegraphics{inelasticity_profile}
  \caption{
    \label{fig:inelasticity_profile}
    Estimates of the normalized proton-proton inelasticity density $G_\mathrm{inel}(b) = d^2\sigma_\mathrm{inel}/d^2b$ at two different LHC beam energies.
    Blue line: Inelasticity density obtained by fitting a parametric form to LHC data at $\sqrts=7$~TeV \cite{Albacete:2016pmp}.
    Orange line: \trento\ predictions at $\sqrts=5.02$~TeV using median parameters from the present study's Bayesian posterior.
  }
\end{figure}

Additional constraints on the proton size and shape are provided by the proton-proton inelasticity density $G_\mathrm{inel}(b) = d^2\sigma_\mathrm{inel}(b)/d^2b$, which measures the proton-proton inelastic collision probability as a function of its impact parameter.
We compare in Fig~\ref{fig:inelasticity_profile} the normalized proton-proton inelasticity density $G_\mathrm{inel}(b) / G_\mathrm{inel}(0)$ at $\sqrts=5.02$~TeV predicted by \trento\ using the present study's posterior median parameters with a calculation that extracted the proton-proton inelasticity density using a parametrization fit to proton-proton differential scattering data at $\sqrts=7$~TeV \cite{Albacete:2016pmp}.
Our Bayesian median estimate (orange line) prefers a larger width for the proton-proton normalized inelasticity density compared to that of reference \cite{Albacete:2016pmp}, and this trend is \emph{opposite} what one would expect based on the difference in beam energy alone.
This suggests that our posterior estimate for the proton radius is somewhat oversized.
Nevertheless, it is fascinating that the present Bayesian estimate is as close as it is, given that the model is fit to quantities which are not typically used to extract the proton-proton inelasticity density.

@@

Second, we obtained a posterior distribution for the model input parameters in Fig.~\ref{fig:posterior} and reported quantitative estimates for these parameters in Table~\ref{tab:post_param}.
Here we summarize our key findings about the model:
\begin{enumerate}[itemsep=0pt, leftmargin=2\parindent]
  \item
    Using \trento\ initial conditions followed by a pre-equilibrium free streaming stage, we find that the initially produced energy density scales like the geometric mean of participant nuclear thickness; see Eq.~\eqref{eq:geometric_mean}.
    Previous Bayesian studies have reported similar scaling for the initially produced \emph{entropy} density \cite{Bernhard:2016tnd, Ke:2016jrd}, although these studies assumed static initialization, an approximation which neglects the initial values of $u^\mu$, $\pi^{\mu\nu}$ and $\Pi$ at the hydrodynamic starting time.
  \item
    We find that nucleon substructure is necessary to simultaneously describe $p$-Pb and Pb-Pb bulk observables.
    However, we observe no strong preference for a specific number of constituents $\nc$ (hot spots) inside the nucleon.
    In particular, we find no evidence to support the specific number $\nc=3$ which is commonly used in the literature.
  \item
    The present Bayesian analysis prefers larger nucleons, $w \sim 1$~fm, in agreement with a similar Bayesian analysis calibrated to Pb-Pb data at $\sqrts=2.76$ and 5.02~TeV using a hybrid model without nucleon substructure \cite{Bernhard:2018hnz}.
    We note that our result is in significant tension with an estimate for the effective nucleon width based on the gluon distribution extracted from HERA data \cite{Rezaeian:2012ji}.
    Our model calculations also predict a broader proton-proton inelasticity density at $\sqrts=5.02$~TeV than supported by the data \cite{Albacete:2016pmp}.
    This suggests that our reconstructed protons are somewhat oversized.
  \item
    We obtain an estimate $\wc=0.52_{-0.20}^{+0.28}$~fm for the Gaussian width of the constituent hot spots inside each nucleon.
    This width is considerably larger than the length scales typically associated with nucleon substructure.
    However, it is natural to wonder if this estimate is oversized, given the aforementioned comments regarding our extracted nucleon width.
  \item
    We report an estimate $\taufs=0.47_{-0.37}^{+0.55}~\fmc$ for the model's pre-equilibrium free streaming time.
    This is significantly shorter than the estimate obtained from a similar Bayesian study in Ref.~\cite{Bernhard:2018hnz}, which found $\taufs=1.16_{-0.25}^{+0.29}~\fmc$.
    It is not clear whether the difference is a result of nucleon substructure or the different observables used to calibrate each analysis.
  \item
    We compare in Fig.~\ref{fig:region_shear_bulk} our estimate for the temperature dependence of the QGP specific shear and bulk viscosities to those of Ref.~\cite{Bernhard:2018hnz}, which performed a Bayesian calibration to Pb-Pb bulk observables at $\sqrts=2.76$ and 5.02~TeV using a physics model without nucleon substructure.
    The two studies are in good agreement, although Ref.~\cite{Bernhard:2018hnz} obtains a more precise estimate for $(\eta/s)(T)$, likely due to the additional beam energies and observables included and an enhanced sensitivity of larger collision systems to the QGP viscosity.
  \item
    We make predictions in Figs.~\ref{fig:obs_map} and \ref{fig:flow_corr} for several quantities which were not included in the model calibration, including the identified yields, transverse energy, symmetric cumulants, and mean $p_T$ fluctuations at $\sqrts=5.02$~TeV.
    Interestingly, our MAP estimate for the Pb-Pb symmetric cumulants at $\sqrts=5.02$~TeV are significantly different than those estimated in Ref.~\cite{Bernhard:2018hnz}.
    This could be a direct (or indirect) result of including nucleon substructure in the model calibration.
\end{enumerate}

@@

Similarly, the results would greatly benefit from additional beam energies and collision systems.
For example, future studies should incorporate proton-proton collisions at top LHC energies, as well as the numerous collisions systems studied at top RHIC energy.
Such data would undoubtedly improve constraints on the model parameters and would enable more stringent tests of the calibrated model predictions.
The RHIC data may also help elucidate the beam-energy dependence of the model parameters which would be worth investigating.
We leave these improvements for future studies.
